---
title: "Gapmindr Analysis"
author: "Sofia Urgoiti Crespo"
format: 
  html:
    code-fold: true
    toc: true
execute:
  echo: true
  message: false
  warning: false
editor: visual
---

```{r}
library(tidyverse)
library(gapminder)
```


1. For each year and continent, summarize average per capita GDP across all countries in the continent. Hint: don't just average country-level values - need to consider country populations!

```{r}
gapminder

summary_gapminder = gapminder %>%
  group_by(continent, year) %>%
  summarize(mean_gdp = mean(gdpPercap)) %>%
  ungroup()
```


2. Plot continent-level GDP values in scatterplot. Bonus, plot the country level values as well with a low alpha to fade them into the background.

```{r}
ggplot(summary_gapminder, aes(x = continent, y = mean(gdpPercap))) +
         geom_point()
       
```


3. Redo the scatterplot but this time log transform the GDP data (in df use mutate or on the plot scale_ly_log10)



4. Based on the plots choose either regular or log transformed data and use a linear regression to estimate annual change in per capita GDP ** or lof(percapitaGDP) accounting for continent.






